---
title: "ClinicalNLP-KCH: MedCAT Baseline Validation Report"
author: "NHS IT Engineer → AI Engineering Transition Project"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
library(ggplot2)
library(dplyr)
library(jsonlite)
library(knitr)
```

## Project Overview

This report evaluates the baseline performance of a MedCAT 2.6.0 clinical NLP pipeline
for SNOMED-CT concept extraction from synthetic NHS discharge summaries.
The pipeline mirrors the CogStack/MedCAT architecture deployed at King's College Hospital.

## Baseline Results
```{r load-data}
results <- fromJSON("../data/baseline_results.json")

overall <- data.frame(
  Metric    = c("Precision", "Recall", "F1 Score"),
  Score     = c(results$overall$precision,
                results$overall$recall,
                results$overall$f1)
)

kable(overall, digits=3, caption="MedCAT Overall Performance")
```

## Per-Summary Performance
```{r per-summary}
per_sum <- results$per_summary %>%
  select(id, precision, recall, f1)

kable(per_sum, digits=3,
      col.names=c("Summary ID","Precision","Recall","F1"),
      caption="Performance by Discharge Summary")
```

## F1 Score by Summary
```{r f1-plot, fig.width=8, fig.height=4}
per_sum %>%
  ggplot(aes(x=id, y=f1, fill=f1)) +
  geom_col() +
  geom_hline(yintercept=results$overall$f1,
             linetype="dashed", colour="red", linewidth=0.8) +
  scale_fill_gradient(low="#f8d7da", high="#0d6efd") +
  annotate("text", x=1, y=results$overall$f1+0.03,
           label=paste("Overall F1:", round(results$overall$f1,3)),
           colour="red", size=3.5) +
  labs(title="MedCAT F1 Score per NHS Discharge Summary",
       subtitle="Dashed line = overall F1. Lower scores indicate missed multi-word terms.",
       x="Summary ID", y="F1 Score") +
  ylim(0, 1.1) +
  theme_minimal() +
  theme(legend.position="none")
```

## Precision vs Recall
```{r pr-plot, fig.width=7, fig.height=5}
per_sum %>%
  ggplot(aes(x=recall, y=precision, label=id)) +
  geom_point(size=4, colour="#0d6efd") +
  geom_text(nudge_y=0.03, size=3) +
  geom_hline(yintercept=results$overall$precision,
             linetype="dashed", colour="orange") +
  geom_vline(xintercept=results$overall$recall,
             linetype="dashed", colour="orange") +
  labs(title="Precision vs Recall by Discharge Summary",
       subtitle="Orange lines = overall averages",
       x="Recall", y="Precision") +
  xlim(0, 1.1) + ylim(0, 1.1) +
  theme_minimal()
```

## Failure Analysis
```{r missed}
missed_df <- data.frame(
  Term = c("acute myocardial infarction",
           "type 2 diabetes",
           "atrial fibrillation",
           "heart failure"),
  SNOMED = c("57054005","44054006","49436004","84114007"),
  Words  = c(3, 3, 2, 2),
  Reason = c("3-word span","3-word span","2-word span","2-word span")
)

kable(missed_df,
      col.names=c("Missed Term","SNOMED Code","Word Count","Failure Reason"),
      caption="Entities Missed by MedCAT Baseline")
```
```{r missed-plot, fig.width=7, fig.height=3}
missed_df %>%
  ggplot(aes(x=reorder(Term, Words), y=Words, fill=Reason)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values=c("2-word span"="#f0ad4e","3-word span"="#d9534f")) +
  labs(title="MedCAT Failure Mode: Multi-word Clinical Terms",
       subtitle="All missed entities are multi-word spans — motivation for LLM fine-tuning",
       x="", y="Word Count") +
  theme_minimal()
```

## Conclusion

MedCAT achieves strong precision (0.923) but reduced recall (0.818) due to consistent
failure on multi-word clinical terms. This motivates the Unsloth QLoRA fine-tuning
approach, where the LLM is trained to capture complex spans that rule-based NER misses.

| Approach | Precision | Recall | F1 | Multi-word Terms |
|----------|-----------|--------|----|-----------------|
| MedCAT baseline | 0.923 | 0.818 | 0.867 | Missed |
| Fine-tuned LLM (target) | TBD | TBD | TBD | Captured |
| Hybrid (production) | High | High | >0.95 | Captured |

The hybrid architecture — MedCAT for code precision, LLM for recall — is the
approach deployed by CogStack at King's College Hospital NHS Foundation Trust.
